---
title: 'Technical Note: <br> Primer on Longitudinal Data Analysis <br> via Linear Mixed-Effects Models'
author: "Stu Field"
date: "`r format(Sys.Date(), '%A %B %d, %Y')`"
transition: "rotate"
output:
  xaringan::moon_reader:
    css: ["css/mtheme_max.css", "css/fonts_mtheme_max.css"]
    self_contained: false
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightLanguage: R
      countIncrementalSlides: false
---



```{r global_options, include = FALSE}
knitr::opts_chunk$set( 
	tidy       = FALSE,     # should output be cleaned up? 
	fig.align  = "left",    # alignment for figures; left, right, center
	cache      = FALSE,     # should the chunk be cached?  
	eval       = TRUE,      # evaluate
	fig.height = 7,         # default fig height
	fig.width  = 7,         # default fig width
	cache.path = "cache/",  # the path for the cached chunks  
	include    = TRUE,      # should the chunk be included as output? 
	echo       = TRUE       # should the chunk be echoed to the pdf? 
)
options(replace.assign = TRUE, digits = 3, width = 175)
library(SomaMixedEffects)
library(ggplot2)
```


# Intro to the Linear Model

.pull-left[
#### Simple Linear Regression

$$\begin{eqnarray}
  y_i      &\sim& \beta_0 + \beta_1 x_i + \epsilon_i \\
  \epsilon &\sim& N(0,\sigma^2_n)
\end{eqnarray}$$

for the $i^{th}$ sample/observation. Thus, we can estimate the
model coefficients via:

$$\begin{equation} 
  \hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i 
\end{equation}$$
]

.pull-right[
#### Extension to Multiple Regression
The linear model is easily expanded to include multiple predictors
(regressors / variables / proteins).

$$\begin{eqnarray}
  y_i &=& \beta_0 x_{0i} + \beta_1 x_{1i} + \dots + \beta_p x_{pi} + \epsilon_i \\
  \epsilon_i &\sim& \text{NID}(0,\sigma^2)
\end{eqnarray}$$

for the $i^{th}$ sample and $p^{th}$ covariate. We typically set
$x_{0i}=1$ so that $\beta_0$ is a constant intercept.
In this form, the only *random-effect* is the error term, $\epsilon_i$. 
]


---

# Intro to the Linear Model

## The Matrix form:

The previous equation can be re-written in matrix form as:
$$\begin{eqnarray}
  \begin{pmatrix}
     y_1 \\ y_2 \\ \vdots \\ y_n
   \end{pmatrix}
    &=&
   \begin{pmatrix}
     x_{11} & x_{12} & \dots & x_{1p} \\
      x_{21} & x_{22} & \dots & x_{2p} \\
     \vdots & \vdots & \ddots & \vdots \\
     x_{n1} & x_{n2} & \dots & x_{np} \\
   \end{pmatrix}
   *
  \begin{pmatrix}
	\beta_1 \\ \beta_2 \\ \vdots \\ \beta_p
  \end{pmatrix}
   +
  \begin{pmatrix}
     \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_n
 \end{pmatrix}
 \\ \nonumber && \\ \nonumber
  \vec y &=& X \vec \beta + \vec \epsilon
\end{eqnarray}$$

where $\vec y$ is a $n\times1$ vector, $X$ is the $n\times p$ data matrix,
$\vec \beta$ is a $p\times1$ vector of fixed effects coefficients for
the $p$ covariates, and $\vec \epsilon$ is the $n\times1$ vector of
observation specific errors.


---


# The Mixed-Effect Model

.pull-left[
- Mixed-effects models fit both **fixed** effects *and* **random** effects
  in the same model. 
- Assume grouping or dependence structure (typically temporal or spatial)
  sampled from a larger population 
    + e.g. plots within a farm, schools within districts, or
      observations within subjects).
- Longitudinal data: 
    + samples from the same individual reflecting serial dependence
- Compound symmetry:
    + specific structure in the variance-covariance matrix (off-diagonal are equal)
    + samples within a subject vary in the same way across subjects
]

--

.pull-right[
- Model must account for this *non-independence* in sample structure. 
- We usually want to fit subject-specific (random) intercepts 
    + parallel linear fits can hit the y-axis independently
]



---

# The Mixed-Effect Model
## Model specification

$$\begin{eqnarray*}
  y_{ij} &=& b_{0i} + \beta_1 x_{ij} + \epsilon_{ij} \\
  b_i &\sim& N(0,\sigma^2_b) \\
	\epsilon &\sim& N(0,\sigma^2_n)
\end{eqnarray*}$$

where:
.pull-left[
- for the $j^{th}$ observation of the $i^{th}$ subject
- $b$ is a parameter treated as a random variable (i.e. $i$ subjects
  sampled from a large, infinite population of potential subjects). 
]

.pull-right[
- Random effects are assumed to vary by group (i.e. subject!) 
+ The ultimate goal is to capture the subject specific variation in
  the coefficients, thus preventing it from being packaged into $\epsilon$!
]


---

# The Mixed-Effect Model
## Syntax in R
- Use either the `nlme` or `lme4` package
- R uses the *Wilkinson* notation in specifying model formulae:
  + Random Intercept
```r
lme(y ~ time * group, random = ~ 1 | pid, data = adat)
```
  + Random Slope
```r
lme(y ~ time * group, random = ~ time | pid, data = adat)
```
  + Random Intercept & Random Slope
```r
lme(y ~ time * group, random = ~ 1 + time | pid, data = adat)
```


---

# The Mixed-Effect Model

## Syntax

| Symbol         | Corresponds to                                   |
| -------------- | :-----------------------------------------------:|
| $y$            | $y_{ij}$                                         |
| $∼ 1$          | random effects; $b_{0i}$                         |
| `pid`          | field containing the dependent groups (subjects) |
| `time * group` | fixed effects (with interaction term)            |
| `time * group` | `time + group + time * group`                    |


---

layout: false
class: inverse, middle, center

# Example 1: Response to Drug


---

# Example 1: Response to Drug

.pull-left[
- Simulate longitudinal data 
  + $20$ subjects, having $3-10$ serial samples each
  + intercept $\beta_0 = 1000$; slope $\beta_1 = 400$ 
  + serial autocorrelation parameter of $0.1$
- The default simulation values can be seen below:

```{r sim_long, warning = FALSE}
# simulate longitudinal data
# createLongData() is a wrapper for simulateLongData()
p <- list(nsubj = 20, beta0 = 1000, beta1 = 400,
          max.obs = 10, r.seed = 99,
          sd.pars = list(sigma = 250), auto.cor = 0.1)
data_response <- createLongData(subject = p)
```
]

.pull-right[
```{r sim_long_cont}
data_response
```
]



---

# Example 1: Response to Drug

.code70[
```{r plot_long_data, echo = TRUE, fig.height = 5.5, fig.width = 10}
y_lab <- bquote("Response ("*y[ij]*")")
data_response |>
  ggplot(aes(x = time, y = yij, group = pid, colour = pid)) +
    geom_line() + ggtitle(bquote(y[ij]~"~"~time)) +
    geom_point(size = 4, shape = 21, fill = "white") +
    ylab(y_lab) + guides(colour = guide_legend(ncol = 1)) +
    facet_wrap(vars(pid)) 
```
]


---

# Example 1: Response to Drug

.code70[
```{r plot_long_data2, fig.height = 6, fig.width = 9}
data_response |>
  ggplot(aes(x = time, y = yij, group = pid, colour = pid)) +
    geom_line() + ggtitle(bquote(y[ij]~"~"~time)) +
    geom_point(size = 4, shape = 21, fill = "white") +
    guides(colour = guide_legend(ncol = 1)) +
    ylab(y_lab) + facet_wrap(vars(Group))
```
]



---

# Example 1: Analysis in R

.pull-left[
The model specification to fit:
$$
\begin{equation}
  y_{ij} = b_{0i} + \beta_1 time_{ij} + \epsilon_{ij},
\end{equation}
$$

again for the $j^{th}$ observation of the $i^{th}$ subject,
where $b_{0i}$ is the subject-specific random *intercept*,
and $\hat{\beta_1}$ is the fixed-effect for *time*.

```{r fit1}
# subject-specific random intercepts with fixed time effect
fit1 <- fit_lme_safely(yij ~ time, random = ~ 1 | pid,
                       data = data_response)
```
]

.pull-right[
.code50[
```{r fit1_cont}
summary(fit1)
```
]
]

From the fixed-effects column in `Value`, the estimates are reasonably
close to the original parameters:
- $\hat{\beta_0}$ = `r summary(fit1)[["tTable"]]["(Intercept)", "Value"]` and
  $\hat{\beta_1}$ = `r summary(fit1)[["tTable"]]["time", "Value"]`
- Both the slope and intercepts differ significantly from zero.


---

layout: false
class: inverse, middle, center

# Example 2: Group Dependent Response (Interaction)


---

# Example 2: Group Dependent Response (Interaction)

.pull-left[
In clinical studies, it is often of interest to understand how the 
treatment of one group of subjects differs from another group of 
subjects (e.g. a control group). In a mixed-model setting, this is
accomplished by adding an interaction term to the model, which sets 
up a conditional response variable, given that a subject belongs to 
a particular group/class.
]

.pull-right[
Next, simulate longitudinal data for 20 subjects, having $3-10$ serial
samples each, intercepts of $\beta_0 = 1000$, slopes of $\beta_1 = 0$ (crtl)
and $\beta_1 = 500$ (treat), and a serial autocorrelation parameter of 0.1.
]

```{r sim_long2, warning = FALSE}
control <- list(nsubj = 20, beta1 = 0, max.obs = 10, r.seed = 9,
                sd.pars = list(sigma = 150), auto.cor = 0.1)

treatment <- list(nsubj = 20, beta1 = 500, max.obs = 10, r.seed = 9,
                  sd.pars = list(sigma = 350), auto.cor = 0.1)

data_group_response <- createLongData(control, treatment)
```



---

# Example 2: `ggplot`
.code70[
```{r plot_long_data3, fig.width = 14, fig.height = 6}
data_group_response |>
  ggplot(aes(x = time, y = yij, group = pid, colour = pid)) +
  geom_line() + ylab(y_lab) +
  geom_point(size = 4, shape = 21, fill = "white") +
  guides(colour = guide_legend(ncol = 2)) +
  facet_wrap(vars(Group)) + ggtitle(bquote(y[ij]~"~"~time))
```
]



---

# Example 2: Model Specification

The model specification to fit:

$$\begin{equation}
  y_{ij} = b_{0i} + \beta_1 time_{ij} + \beta_2 group_i + \beta_3 (group_i \times time_{ij}) + \epsilon_{ij},
\end{equation}$$

.pull-left[
again for the $j^{th}$ observation of the $i^{th}$ subject, $b_{0i}$ is 
still the subject-specific random intercept and $\beta_1$ is still the
fixed-effect for *time*.
]

.pull-right[
However, now there is a fixed-effect for *group*
and an interaction term, $\beta_3$ term, that is multiplied by a **dummy**
variable such that:
$$\begin{equation}
  G = \begin{cases}
  0, & control \\
  1, & treatment
  \end{cases}
\end{equation}$$
]

giving,

$$\begin{equation}
	y_{ij} = b_{0i} + \beta_1 time_{ij} + \beta_2 group_i + \beta_3 (group_i \times time_{ij}) * G + \epsilon_{ij}
\end{equation}$$


---

# Example 2: Analysis in R

.pull-left[
- It is often a good idea to check the variation on the coefficients of
  subject-specific linear fits of the data.
]

.pull-right[
- The `SomaMixedEffects::lmeDiagnostic()` function is a wrapper around
  `nlme::lmList()` and creates linear models for each subject *i*.
]
.code70[
```{r fit2, fig.width = 14, fig.height = 5}
fit2 <- fit_lme_safely(yij ~ time * Group, random = ~ 1 | pid, data = data_group_response)
lmeDiagnostic(fit2)     # plot intervals of coefficients
```
]



---

# Example 2: Summary

.pull-left[
- Notice that the subject-specific offsets, $\beta_0$, are approximately
  centered about 1000 (RFU) for both treatment and control (as they should
  be; we set them to the same intercept)
- A vertical line up from $1000$ would cross the confidence interval 
  for most subjects.
- The more variable the estimates (and intervals) of $\beta_0$ seen in the
  left panel, the greater the improvement in model fit can be found in
  fitting subject-specific offsets in a mixed-effects model framework.
]

--

.pull-right[
- Secondly, as can be expected, the slope coefficient, $\beta_1$, is 
  vastly different between treated and control. 
- Depending on the statistical question, subject-specific slopes may be
  desired (but typically not for longitudinal time-series data 
  in a clinical setting).
]


---

# Example 2: Analysis in R

.pull-left[
.code50[
```{r fit_summary}
summary(fit2)
```
]
]

.pull-right[
From the fixed-effects column in `Value`, the estimates are
reasonably close to the original parameter:
$\hat{\beta_0}=$ `r summary(fit2)[["tTable"]]["(Intercept)", "Value"]` and
$\hat{\beta_1}=$ `r summary(fit2)[["tTable"]]["time", "Value"]`.
The significant interaction indicates that the entire fixed-effect
for *time* occurs in the *treatment* group; in this case the *control*
group does not vary with time.
]


---

layout: false
class: inverse, middle, center

# Example 3: Group Dependent Response (2 Slopes)


---

# Example 3: Group Dependent Response (2 Slopes)

To illustrate how the model output is constructed, simulate identical data
to above, but give the control group a slope of $\beta_1 = 250$.
This results in a significant fixed-effect for time *and* an
significant interaction; the additional effect of being in the *treatment*
group. Relate the values back to the original Equation~**X** above to see how
the terms are partitioned.


```{r sim_long3, warning = FALSE}
# simulate longitudinal data with group-specific response
p1 <- list(nsubj = 20, beta1 = 250, max.obs = 10, r.seed = 5,
           sd.pars = list(sigma = 150), auto.cor = 0.1)

p2 <- list(nsubj = 20, beta1 = 500, max.obs = 10, r.seed = 5,
           sd.pars = list(sigma = 350), auto.cor = 0.1)

data_group_response2 <- createLongData(control = p1, treatment = p2)
```


---

# Example 3: Group Dependent Response (2 Slopes)
.code70[
```{r plot_long_data4, fig.width = 14, fig.height = 6}
data_group_response2 |>
  ggplot(aes(x = time, y = yij, group = pid, colour = pid)) +
  geom_line() + ylab(y_lab) +
  geom_point(size = 4, shape = 21, fill = "white") +
  guides(colour = guide_legend(ncol = 2)) +
  facet_wrap(vars(Group)) + ggtitle(bquote(y[ij]~"~"~time))
```
]


---

# Example 3: Analysis in R
.pull-left[
.code50[
```{r fit3}
fit3 <- fit_lme_safely(yij ~ time * Group, random = ~ 1 | pid,
                       data = data_group_response2)
summary(fit3)
```
]
]

.pull-right[
Notice that the
$\hat{\beta_0}=$ `r summary(fit3)[["tTable"]]["(Intercept)", "Value"]`
has not changed, as we haven not changed any baseline values, but now half of 
the effect has moved into the *time* effect,
$\hat{\beta_1}=$ `r summary(fit3)[["tTable"]]["time", "Value"]`, and away 
from the interaction coefficient,
$\hat{\beta_3}=$ `r summary(fit3)[["tTable"]]["time:Grouptreatment", "Value"]`,
which was carrying the entire slope effect for the treatment group. 
]



---

# Xaringan and Session Info

.pull-left[
This is a [xaringan](https://github.com/yihui/xaringan) 
(v`r utils::packageVersion("xaringan")`) presentation
created on `r Sys.Date()`.
Written by [Rmarkdown](https://github.com/rstudio/rmarkdown)
(v`r utils::packageVersion("rmarkdown")`) and `r R.version$version.string`.
]
.pull-right[
]

.code50[
```{r, echo = FALSE}
utils::sessionInfo()
```
]
